---
title: "moreTCGAsurvival"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(survival)
library(survminer)
library(cgdsr)
```

##More Survival analysis of TCGA samples
This time we have more samples. We will extrapolate information from the Tumor/healthy tissue matched samples (we can caluculate delta psis) to samples which are not matched. 
Matt gave me a list of significant delta psis in the direction that a cancer globally leans (each cancer can either lean positive or negative globally). These genes and their average (?) deltapsi are in "leaningdpsis".

```{r, read in data}
##delta psis are Tumor - Normal
leaningdpsis <- as_tibble(read.table("data/TCGApsis/leaningdpsis.txt", header = TRUE))

BLCA_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.BLCA.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
BRCA_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.BRCA.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
COAD_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.COAD.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
HNSC_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.HNSC.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
KIRC_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.KIRC.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
KIRP_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.KIRP.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
LIHC_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.LIHC.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
LUAD_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.LUAD.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
LUSC_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.LUSC.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
PCPG_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.PCPG.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
PRAD_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.PRAD.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
STAD_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.STAD.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))
THCA_psis <- as_tibble(read.table("data/TCGApsis/LABRAT.THCA.psis", header = TRUE)) %>% gather(-Gene, -genetype, key = patient, value = psi) %>% na.omit() %>% mutate(Gene = unlist(lapply(Gene, function(x) unlist(str_split(x, "[.]"))[1])))

#psi_list <- list(BLCA_psis, BRCA_psis, COAD_psis, HNSC_psis, KIRC_psis, KIRP_psis, LIHC_psis, LUAD_psis, LUSC_psis, PRAD_psis, STAD_psis, THCA_psis)
#names(psi_list) <- c('BLCA', 'BRCA', 'COAD', 'HNSC', 'KIRC', 'KIRP', 'LIHC', 'LUAD', 'LUSC', 'PRAD', 'STAD', 'THCA')

#TCGA_psis <- bind_rows(psi_list, .id = "cancer")

#old m_dat from Austin:
#m_dat <- as_tibble(read.csv("data/TCGApsis/tcga_clinical_meta.tsv", header = TRUE, sep = "\t")) %>% 
#  mutate(patient = str_replace_all(Case_ID, pattern = "-", replace = "."))

cancernames <- c('BLCA', 'BRCA', 'COAD', 'HNSC', 'KIRC', 'KIRP', 'LIHC', 'LUAD', 'LUSC', 'PRAD', 'STAD', 'THCA')
case_list <- c("blca_tcga_all", "brca_tcga_all", "coadread_tcga_all", "hnsc_tcga_all", "kirc_tcga_all", "kirp_tcga_all", "lihc_tcga_all", "luad_tcga_all", "lusc_tcga_all", "prad_tcga_all", "stad_tcga_all", "thca_tcga_all")

#####PCPG DOESNT HAVE OS MONTHS?????


mycgds = CGDS("https://www.cbioportal.org/")

x <- 1:12
CD <- lapply(x, function(x) select(as_tibble(getClinicalData(mycgds, case_list[x]),rownames="patient"),patient,OS_MONTHS,OS_STATUS))

m_dat <- CD %>%
  bind_rows() %>% 
  mutate(patient = substr(patient,0,nchar(patient)-3),
         OS_STATUS = as.numeric(substr(OS_STATUS, 0,1)))


LUSC_leandir <- "negative"
BLCA_leandir <- "negative"
LUAD_leandir <- "negative"
HNSC_leandir <- "negative"
LIHC_leandir <- "negative"
BRCA_leandir <- "negative"
COAD_leandir <- "negative"
PRAD_leandir <- "negative"
KIRP_leandir <- "positive"
THYM_leandir <- "positive"
PCPG_leandir <- "positive"
THCA_leandir <- "positive"
KIRC_leandir <- "positive"
STAD_leandir <- "positive"

```

##categorize patients for each cancer type
patients will be categorized as extremely affected or not via their psi values for the genes of interest. 
Considering only psi values for the genes of interest, median or mean psis are calculated for each patient.
These values are then ordered from low to high and patients are split into extreme halves, thirds or quartiles.


```{r, plot_fx, fig.height=7, warning=FALSE}
#old m_dat fx
cancer_surv_plot <- function(cancer, rank_value, group, title){
  
  cancer_psis <- eval(parse(text = paste(cancer,"_psis", sep = "")))
  
  cancer_genes <- leaningdpsis %>% filter(tumor == cancer) %>% pull(Gene)
  cancer_ranks <- cancer_psis %>% 
    filter(Gene %in% cancer_genes) %>% 
    group_by(patient) %>% 
    summarize(n=n(), 
              mean_psi = mean(psi, na.rm = TRUE),
              med_psi = median(psi, na.rm = TRUE),
              short = length(which(psi < 0.5)) / n * 100,
              long = length(which(psi > 0.5)) / n * 100, .groups = "rowwise") %>% 
    filter(n > 0.75*length(cancer_genes))  %>% 
    left_join(., dplyr::select(m_dat, patient, vital_status, days_to_death)) %>%
    mutate(days = ifelse(days_to_death == "'--", NA, as.numeric(days_to_death)),
           event = ifelse(vital_status == "Alive", 1, ifelse(vital_status == "Dead", 2, 3))) %>% 
    filter(event != 3) %>% #this excludes events that are "Not Reported"
    arrange(eval(parse(text = rank_value))) %>% 
    ungroup() %>% 
    mutate(rank = c(1:nrow(.)),
           halfs = ifelse(rank > round(nrow(.)/2), "high", "low"),
           thirds = ifelse(rank < round(nrow(.)/3), "low", ifelse(rank > round(nrow(.)*(2/3)), "high", "middle")),
           ExtQs = ifelse(rank < round(nrow(.)/4), "low", ifelse(rank > round(nrow(.)*(3/4)), "high", "middle"))) %>% 
    filter(eval(parse(text = group) != "middle"))
  
  sfit2 <- surv_fit(Surv(cancer_ranks$days, cancer_ranks$event)~eval(parse(text = group)), data = cancer_ranks)
  ggsurvplot(sfit2, pval = TRUE, pval.method = TRUE, risk.table = TRUE, legend.labs = substr(names(sfit2$strata),27,nchar(names(sfit2$strata)))) + labs(title = paste(cancer, rank_value, "by", group, sep = " "), subtitle = paste("Deltapsis lean", eval(parse(text = paste(cancer,"_leandir", sep = ""))), sep = " "))
  
}

#cgdsr m_dat fx
cancer_surv_plot <- function(cancer, rank_value, group){
  
  cancer_psis <- eval(parse(text = paste(cancer,"_psis", sep = "")))
  
  cancer_genes <- leaningdpsis %>% filter(tumor == cancer) %>% pull(Gene)
  cancer_ranks <- cancer_psis %>% 
    filter(Gene %in% cancer_genes) %>% 
    group_by(patient) %>% 
    summarize(n=n(), 
              mean_psi = mean(psi, na.rm = TRUE),
              med_psi = median(psi, na.rm = TRUE), .groups = "rowwise") %>% 
    filter(n > 0.75*length(cancer_genes)) %>% 
    left_join(., m_dat) %>% 
    na.omit() %>% 
    arrange(eval(parse(text = rank_value))) %>% 
    ungroup() %>% 
    mutate(rank = c(1:nrow(.)),
           halfs = ifelse(rank > round(nrow(.)/2), "high", "low"),
           thirds = ifelse(rank < round(nrow(.)/3), "low", ifelse(rank > round(nrow(.)*(2/3)), "high", "middle")),
           ExtQs = ifelse(rank < round(nrow(.)/4), "low", ifelse(rank > round(nrow(.)*(3/4)), "high", "middle"))) %>% 
    filter(eval(parse(text = group)) != "middle")
  
  sfit2 <- surv_fit(Surv(cancer_ranks$OS_MONTHS, cancer_ranks$OS_STATUS)~eval(parse(text = group)), data = cancer_ranks)
  
  ggsurvplot(sfit2, pval = TRUE, 
             pval.method = TRUE, 
             risk.table = "nrisk_cumcensor",
             legend.title = "Psi",
             legend.labs = substr(names(sfit2$strata), 27, nchar(names(sfit2$strata))),
             xlab = "Months",
             title = paste(cancer, rank_value, "by", group, "\n", "Deltapsis lean", eval(parse(text = paste(cancer,"_leandir", sep = ""))), sep = " "))
  
}

```

##Expectations
Cancer types that have delta psis that lean Negative (more commmon) are expected to have worse outcome when psis are LOW
Negative == High psis on top, low psis bottom left

Cancer types that have delta psis that lean Positive are expected to have worse outcome when psis are HIGH
Positive == Low psis on top, high psis bottom left

##Median Psi Values in extreme upper and lower quartilies

```{r, med_psi_qs, fig.height=7, warning=FALSE}
lapply(x, function(x) cancer_surv_plot(cancernames[x], "med_psi", "ExtQs"))

```

```{r, pretty plots}
HNSC_genes <- leaningdpsis %>% filter(tumor == "HNSC") %>% pull(Gene)

  HNSC_ranks <- HNSC_psis %>% 
    filter(Gene %in% HNSC_genes) %>% 
    group_by(patient) %>% 
    summarize(n=n(), 
              med_psi = median(psi, na.rm = TRUE), .groups = "rowwise") %>% 
    filter(n > 0.75*length(HNSC_genes)) %>% 
    left_join(., m_dat) %>% 
    na.omit() %>% 
    arrange(med_psi) %>% 
    ungroup() %>% 
    mutate(rank = c(1:nrow(.)),
           ExtQs = ifelse(rank <= round(nrow(.)/4), "low", ifelse(rank > round(nrow(.)*(3/4)), "high", "middle"))) %>% 
    filter(ExtQs != "middle")
  
  sfit_HNSC <- surv_fit(Surv(HNSC_ranks$OS_MONTHS, HNSC_ranks$OS_STATUS)~ExtQs, data = HNSC_ranks)
  
ggsurvplot(sfit_HNSC, 
             pval = TRUE, 
             pval.method = TRUE, 
             palette = c("#FCA285", "#85BCDB"),
             size = 1.5,
             censor.shape = 124,
             legend.title = "median psi",
             legend.labs = paste(substr(names(sfit_HNSC$strata), 7, nchar(names(sfit_HNSC$strata))), " (N=", sfit_HNSC$n, ")", sep = ""),
             font.legend = 16,
             xlab = "Time (Months)",
             title = paste("HNSC", sep = " "),
             font.main = 20,
             xlim = c(0,80),
             break.time.by = 20,
             ylim = c(0.3,1),
             pval.coord = c(0, 0.53),
             pval.method.coord = c(0, 0.6)) 
  
  
  
  
  
  KIRC_genes <- leaningdpsis %>% filter(tumor == "KIRC") %>% pull(Gene)

  KIRC_ranks <- KIRC_psis %>% 
    filter(Gene %in% KIRC_genes) %>% 
    group_by(patient) %>% 
    summarize(n=n(), 
              med_psi = median(psi, na.rm = TRUE), .groups = "rowwise") %>% 
    filter(n > 0.75*length(KIRC_genes)) %>% 
    left_join(., m_dat) %>% 
    na.omit() %>% 
    arrange(med_psi) %>% 
    ungroup() %>% 
    mutate(rank = c(1:nrow(.)),
           ExtQs = ifelse(rank <= round(nrow(.)/4), "low", ifelse(rank > round(nrow(.)*(3/4)), "high", "middle"))) %>% 
    filter(ExtQs != "middle")
  
  sfit_KIRC <- surv_fit(Surv(KIRC_ranks$OS_MONTHS, KIRC_ranks$OS_STATUS)~ExtQs, data = KIRC_ranks)
  
  ggsurvplot(sfit_KIRC, 
             pval = TRUE, 
             pval.method = TRUE, 
             palette = c("#FCA285", "#85BCDB"),
             size = 1.5,
             censor.shape = 124,
             legend.title = "median psi",
             legend.labs = paste(substr(names(sfit_KIRC$strata), 7, nchar(names(sfit_KIRC$strata))), " (N=", sfit_KIRC$n, ")", sep = ""),
             font.legend = 16,
             xlab = "Time (Months)",
             title = paste("KIRC", sep = " "),
             font.main = 20,
             xlim = c(0,100),
             break.time.by = 25,
             ylim = c(0.3,1),
             pval.coord = c(0, 0.58),
             pval.method.coord = c(0, 0.65))
```

##Mean Psi Values in extreme upper and lower quartiles

```{r, mean_psi_qs, fig.height=7, warning=FALSE}
lapply(x, function(x) cancer_surv_plot(cancernames[x], "mean_psi", "ExtQs"))

```

##Median Psi Values in upper and lower halves

```{r, med_psi_halves, fig.height=7, warning=FALSE}
lapply(x, function(x) cancer_surv_plot(cancernames[x], "med_psi", "halfs"))

```

##Mean psi values by upper and lower halves

```{r, mean_psi_halves, fig.height=7, warning=FALSE}
lapply(x, function(x) cancer_surv_plot(cancernames[x], "mean_psi", "halfs"))

```

##Median psi values by upper and lower thirds

```{r, med_psi_thirds, fig.height=7, warning=FALSE}
lapply(x, function(x) cancer_surv_plot(cancernames[x], "med_psi", "thirds"))

```

##Mean Psi Values in extreme upper and lower thirds

```{r, mean_psi_thirds, fig.height=7, warning=FALSE}
lapply(x, function(x) cancer_surv_plot(cancernames[x], "mean_psi", "thirds"))

```
