---
title: "TCGA Clinical Data Anlaysis"
author: "Rae G"
date: "February 28, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cgdsr)
library(tidyverse)
library(survival)
library(survminer)
library(devtools)


```



## The Data
Alternative polyadenylation data for this analysis was provided by Matt Taliaferro. Psi values were calculated by LABRAT then compared between matched tumor and healthy samples from TCGA. The reported delta psi values are all significant, it is expected that greater delta psi values (in either direction) will correlate with worse survival.

## Survival Analysis in General
Survival analyses allow analysis of events occuring over time without assuming constant rates. The Survival Function: $S(t) = Pr(T>t)$ where S is some probability over time (t) whch is equal to the probability of survival where time of death (T) has not yet occured or is less than time (t). 
Clinical data from TCGA records OS_MONTHS or months since diagnosis and OS_STATUS, alive or deceased. From these two values it can be determined how long until an event (death) has occured or not.


```{r, psi data}
psis <- read.table("TCGApsis2.0.txt", header = TRUE) 
psis_tidy <- psis %>% as_tibble() %>% 
  group_by(tumorID, sample) %>% 
  summarize(med_psi = median(value, na.rm = TRUE)) %>% 
  mutate(sample = as.character(sample),
         sample = substr(sample, 6, nchar(sample)-5),
         sample = gsub("_", ".", sample),
         sample = paste("TCGA.", sample, sep = ""),
         abs_psi = abs(med_psi))

```

```{r, case lists}
mycgds = CGDS("https://www.cbioportal.org/")
CS <- getCancerStudies(mycgds) %>% as_tibble()

cancer_studies <- psis %>% 
  select(tumorID) %>%
  unique() %>% 
  unlist() %>% 
  as.character()  %>%
  tolower() %>%
  paste("tcga", sep = "_") %>% 
  .[c(1:4,6:15,17:20)] %>% 
  c(., "coadread_tcga")

psis_order <- psis_tidy %>% 
  ungroup() %>% 
  mutate(tumorID = as.character(tumorID)) %>% 
  group_by(tumorID) %>%
  summarize(n = n()) %>% 
  arrange(desc(n))

x <- 1:19
case_lists <- lapply(x, function(x) getCaseLists(mycgds, cancer_studies[x])) %>% 
    do.call("rbind", .) %>% 
    as_tibble() %>%
    filter(case_list_name == "All samples") %>% 
    select(case_list_id) %>% 
    mutate(tumorID = toupper(substr(case_list_id, 0, 4))) %>% 
    inner_join(psis_order, by = "tumorID") %>% 
    arrange(desc(n)) %>% 
    select(case_list_id) %>% 
    unlist() %>% 
    as.character()

```

```{r, }
printGGlist <- function(gglist){
  for(gg in gglist){
    print(gg)
  }
}  
  
```

```{r, TCGA_survival function}
TCGA_survival <- function(case_name) {
  CD <- getClinicalData(mycgds, case_name) 
  CD <- CD %>% 
    as_tibble() %>% 
    select(OS_MONTHS, OS_STATUS) %>% 
    mutate(sample = rownames(CD), sample = substr(sample,0,nchar(sample)-3))

  psi_surv <<- inner_join(psis_tidy, CD, by = "sample") %>%
    as_tibble() %>% 
    mutate(STATUS = ifelse(OS_STATUS == "LIVING", 1, 2)) %>% 
    arrange(desc(med_psi)) %>% 
    rowid_to_column(var = "med_psi_id") %>% 
    mutate(posneg2 = ifelse(med_psi > 0, "pos", "neg"),
           posneg3 = ifelse(med_psi_id < nrow(.)*(1/3)+1, "pos", ifelse(med_psi_id < nrow(.)*(2/3)+1, "zero", "neg")),
           posnegExt = ifelse(med_psi_id < nrow(.)*(1/4)+1, "pos", ifelse(med_psi_id < nrow(.)*(3/4)+1, NA, "neg")))
  
  psi_surv_abs <<- psi_surv %>% 
    arrange(desc(abs_psi)) %>%
    rowid_to_column(var = "abs_psi_id") %>% 
    mutate(dysreg2 = ifelse(abs_psi_id < nrow(.)/2+1, "yes", "no"),
           dysreg3 = ifelse(abs_psi_id < nrow(.)*(1/3)+1, "yes", ifelse(abs_psi_id < nrow(.)*(2/3)+1, "maybe", "no")),
           dysregExt = ifelse(abs_psi_id < nrow(.)*(1/4)+1, "yes", ifelse(abs_psi_id < nrow(.)*(3/4)+1, NA, "no")))
  

  sfit2 <- surv_fit(Surv(psi_surv$OS_MONTHS, psi_surv$STATUS)~posneg2, data = psi_surv)
  plot1 <- ggsurvplot(sfit2, pval = TRUE, pval.method = TRUE, risk.table = TRUE) + ggtitle(paste(case_name, " Pos/Neg")) 

  sfit3 <- surv_fit(Surv(psi_surv$OS_MONTHS, psi_surv$STATUS)~posneg3, data = psi_surv)
  plot2 <- ggsurvplot(sfit3, risk.table = TRUE) + ggtitle(paste(case_name, " Pos/Zero/Neg"))
  
  sfitExt <- surv_fit(Surv(psi_surv$OS_MONTHS, psi_surv$STATUS)~posnegExt, data = psi_surv)
  plot3 <- ggsurvplot(sfitExt, pval = TRUE, pval.method = TRUE, risk.table = TRUE) + ggtitle(paste(case_name, " Extreme Pos/Neg"))
  
  sfit2_abs <- surv_fit(Surv(psi_surv_abs$OS_MONTHS, psi_surv_abs$STATUS)~dysreg2, data = psi_surv_abs)
  plot4 <- ggsurvplot(sfit2_abs, pval = TRUE, pval.method = TRUE, risk.table = TRUE) + ggtitle(paste(case_name, " Halves"))

  sfit3_abs <- surv_fit(Surv(psi_surv_abs$OS_MONTHS, psi_surv_abs$STATUS)~dysreg3, data = psi_surv_abs)
  plot5 <- ggsurvplot(sfit3_abs, risk.table = TRUE) + ggtitle(paste(case_name, " Thirds"))
  
  sfitExt_abs <- surv_fit(Surv(psi_surv_abs$OS_MONTHS, psi_surv_abs$STATUS)~dysregExt, data = psi_surv_abs)
  plot6 <- ggsurvplot(sfitExt_abs, pval = TRUE, pval.method = TRUE, risk.table = TRUE) + ggtitle(paste(case_name, " Extreme Quartiles"))
  
  gg <- list(plot1, plot2, plot3, plot4, plot5, plot6)
  printGGlist(gg)
}


```

```{r, message = FALSE}
x <- c(1:10)
lapply(x, function(x) TCGA_survival(case_lists[x]))

```


